---
title: "Advanced Usage: Working with Complex Scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Usage: Working with Complex Scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette covers advanced usage scenarios for the cortar package, including working with multiple transcripts, handling family data, and customizing coverage thresholds.

## Working with Specific Transcripts

### Specifying Transcript Versions

You can specify exact RefSeq transcript versions in your sample file:

```{r eval=FALSE}
library(cortar)

# Create sample file with specific transcripts
sample_data <- data.frame(
  sampleID = c("patient1", "control1"),
  familyID = c(1, 2),
  sampletype = c("test", ""),
  genes = c("DMD", "DMD"),
  transcript = c("NM_004006.3", "NM_004006.3"),
  bamfile = c("/path/to/patient1.bam", "/path/to/control1.bam"),
  coverage = c("het", "het")
)

write.table(sample_data, "samples.tsv", sep = "\t", 
            row.names = FALSE, quote = FALSE)

# Run cortar with specific transcript
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "output"
)
```

### Using Canonical Transcripts

Leave the transcript column blank to use the canonical RefSeq transcript:

```{r eval=FALSE}
sample_data <- data.frame(
  sampleID = c("patient1"),
  familyID = c(1),
  sampletype = c("test"),
  genes = c("DMD"),
  transcript = c(""),  # Will use canonical transcript
  bamfile = c("/path/to/patient1.bam"),
  coverage = c("het")
)
```

## Handling Family Data

### Preventing Intra-Family Comparisons

The familyID column ensures that related samples are not compared to each other:

```{r eval=FALSE}
# Family trio: proband, mother, father
family_data <- data.frame(
  sampleID = c("proband_F1", "mother_F1", "father_F1"),
  familyID = c(1, 1, 1),  # Same family
  sampletype = c("test", "", ""),
  genes = c("NF1", "NF1", "NF1"),
  transcript = c("", "", ""),
  bamfile = c("/path/to/proband.bam", 
              "/path/to/mother.bam", 
              "/path/to/father.bam"),
  coverage = c("het", "het", "het")
)

# Proband will not be compared to mother or father
# But will be compared to unrelated controls
```

### Multiple Families

```{r eval=FALSE}
multi_family_data <- data.frame(
  sampleID = c("proband_F1", "mother_F1", 
               "proband_F2", "father_F2",
               "control_unrelated"),
  familyID = c(1, 1, 2, 2, 3),
  sampletype = c("test", "", "test", "", ""),
  genes = c("DMD", "DMD", "DMD", "DMD", "DMD"),
  transcript = c("", "", "", "", ""),
  bamfile = c("/path/to/F1_proband.bam",
              "/path/to/F1_mother.bam",
              "/path/to/F2_proband.bam",
              "/path/to/F2_father.bam",
              "/path/to/control.bam"),
  coverage = c("het", "het", "het", "het", "het")
)
```

## Custom Coverage Thresholds

### Standard Coverage Types

Cortar provides standard coverage thresholds:

- **het** (heterozygous): 60 reads
- **hom** (homozygous): 30 reads
- **hemi** (hemizygous): 30 reads

### Custom Numeric Thresholds

You can specify custom coverage thresholds:

```{r eval=FALSE}
custom_coverage_data <- data.frame(
  sampleID = c("low_coverage", "high_coverage", "standard"),
  familyID = c(1, 2, 3),
  sampletype = c("test", "test", "test"),
  genes = c("DMD", "DMD", "DMD"),
  transcript = c("", "", ""),
  bamfile = c("/path/to/low.bam", 
              "/path/to/high.bam", 
              "/path/to/standard.bam"),
  coverage = c("20", "100", "het")  # Custom numeric or standard
)
```

## Panel Analysis

### Gene Panel Setup

Panel mode is ideal for analyzing multiple genes across many samples:

```{r eval=FALSE}
# Create sample file for panel analysis
panel_data <- data.frame(
  sampleID = paste0("sample_", 1:10),
  familyID = 1:10,
  sampletype = rep(c("test", ""), each = 5),
  genes = rep("", 10),  # Leave blank for panel mode
  transcript = rep("", 10),
  bamfile = paste0("/path/to/sample_", 1:10, ".bam"),
  coverage = rep("het", 10)
)

write.table(panel_data, "panel_samples.tsv", 
            sep = "\t", row.names = FALSE, quote = FALSE)

# Run with gene panel
cortar(
  file = "panel_samples.tsv",
  mode = "panel",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "panel_output",
  genelist = c(
    "DMD", "TTN", "CFTR", "NF1", "COL2A1",
    "FBN1", "LMNA", "MYH7", "MYBPC3", "SCN5A"
  )
)
```

## Research Mode

### Exploratory Analysis

Research mode provides less stringent filtering for exploratory analyses:

```{r eval=FALSE}
cortar(
  file = "research_samples.tsv",
  mode = "research",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "research_output",
  genelist = c("GENE1", "GENE2", "GENE3")
)
```

**Key Differences from Default/Panel Mode:**
- Different filtering criteria for splicing events
- May report more events (including lower confidence)
- Useful for hypothesis generation

## Batch Processing

### Organizing Multiple Analyses

```{r eval=FALSE}
# Directory structure:
# batch_analysis/
#   ├── cohort1_samples.tsv
#   ├── cohort2_samples.tsv
#   └── cohort3_samples.tsv

cortar_batch(
  folder = "batch_analysis",
  pattern = "*_samples.tsv",
  mode = "panel",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "batch_output",
  genelist = c("DMD", "TTN", "NF1")
)

# Output will be organized by cohort
```

## Working with Different Genome Assemblies

### hg38 (Default)

```{r eval=FALSE}
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "output_hg38"
)
```

### hg19

```{r eval=FALSE}
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg19",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "output_hg19"
)
```

### 1000 Genomes Annotation

```{r eval=FALSE}
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "1000genomes",
  paired = TRUE,
  stranded = 2,
  output_dir = "output_1000g"
)
```

## Handling Different RNA-seq Protocols

### Unstranded RNA-seq

```{r eval=FALSE}
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 0,  # Unstranded
  output_dir = "output"
)
```

### Forward Stranded

```{r eval=FALSE}
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 1,  # Forward stranded
  output_dir = "output"
)
```

### Reverse Stranded (Most Common)

```{r eval=FALSE}
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,  # Reverse stranded
  output_dir = "output"
)
```

### Single-End Sequencing

```{r eval=FALSE}
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = FALSE,  # Single-end
  stranded = 2,
  output_dir = "output"
)
```

## Adding Custom Prefixes

### Organizing Output Files

```{r eval=FALSE}
# Add prefix to distinguish different analyses
cortar(
  file = "samples.tsv",
  mode = "panel",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "output",
  genelist = c("DMD", "TTN"),
  prefix = "cohort1_neuromuscular_"
)

# Output files will be named:
# cohort1_neuromuscular_report.xlsx
# cohort1_neuromuscular_report.pdf
```

## Debug Mode for Troubleshooting

### Enabling Debug Output

```{r eval=FALSE}
cortar(
  file = "samples.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "output",
  debug = TRUE
)
```

**Debug mode creates intermediate files:**
- `0_samplefile.tsv`: Processed sample file
- `1_tx_extraction.tsv`: Extracted transcripts
- `2_gene_to_GRange.tsv`: Gene coordinates
- `3_introns_to_GRange_*.tsv`: Intron coordinates
- `4_introns_other_tx_to_GRange.tsv`: Alternative transcript introns
- `5_introns_jx_to_GRange_*.tsv`: Junction coordinates
- `6_combined_sj.tsv`: Combined junction data

These files help diagnose issues with gene selection, coordinate extraction, or read counting.

## Best Practices

### Sample File Organization

1. **Use clear, unique sample IDs**
2. **Assign family IDs consistently**
3. **Verify BAM file paths before running**
4. **Check genome assembly matches BAM alignment**
5. **Use appropriate coverage thresholds**

### Performance Optimization

1. **Use BAM index files** (.bai) for faster access
2. **Process samples in batches** rather than individually
3. **Consider subsetting** BAM files for faster iteration
4. **Use debug mode** only when troubleshooting

### Quality Control

1. **Check sequencing depth** before analysis
2. **Verify strand orientation** of your library
3. **Confirm transcript versions** match your research question
4. **Review control samples** for appropriate comparisons

## Session Info

```{r eval=FALSE}
sessionInfo()
```
