---
title: "Getting Started with cortar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cortar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The **cortar** package extracts reads at the exon-intron junction from RNA-seq data and reports the proportion of splicing events across each intron. By comparing these values to controls, deviations from normal splicing can be characterized in test samples.

This vignette will guide you through the basic usage of cortar for analyzing RNA splicing patterns.

## Installation

You can install cortar from GitHub with:

```{r eval=FALSE}
# Install BiocManager if not already installed
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Install Bioconductor dependencies
BiocManager::install(c(
  "GenomicRanges",
  "GenomicFeatures",
  "GenomicAlignments",
  "BSgenome.Hsapiens.UCSC.hg38",
  "BSgenome.Hsapiens.UCSC.hg19",
  "BSgenome.Hsapiens.1000genomes.hs37d5"
))

# Install cortar from GitHub
devtools::install_github("kidsneuro-lab/RNA_splice_tool")
```

## Quick Start

```{r eval=FALSE}
library(cortar)

# Run test dataset
run_cortar_test()

# Check output directory for reports
# Default location: <directory>/cortar_test/output
```

## Preparing Your Data

### Creating a Sample File

Before running cortar, you need to create a tab-separated sample file containing information about your samples. The file should have the following columns:

- **sampleID**: A unique identifier for the sample
- **familyID**: A unique identifier for related samples (prevents comparison between related samples)
- **sampletype**: Whether analysis is desired for this sample ("test" or leave blank for controls)
- **genes**: Gene symbol for the gene under investigation (can be blank for panel/research mode)
- **transcript**: Transcript under investigation (blank will default to canonical RefSeq transcript)
- **bamfile**: Absolute or relative file path to the BAM file
- **coverage**: Expected coverage type ("het", "hom", "hemi", or custom numeric value)

Example sample file:

```
sampleID	familyID	sampletype	genes	transcript	bamfile	coverage
proband_1	1	test	DMD	NM_004006.2	/path/to/proband_1.bam	het
control_1	2		DMD	NM_004006.2	/path/to/control_1.bam	het
proband_2	3	test	TTN	NM_001267550.2	/path/to/proband_2.bam	hom
control_2	4		TTN	NM_001267550.2	/path/to/control_2.bam	hom
```

**Important Notes:**
- By default, cortar will not compare samples with the same gene/transcript to avoid comparing samples where the same gene is affected
- For control samples, use a valid gene name and transcript number
- BAM files must be aligned to the same genome assembly (hg38 or hg19)

## Running Cortar

### Default Mode

In default mode, each sample can specify its own gene of interest in the sample file:

```{r eval=FALSE}
cortar(
  file = "path/to/samplefile.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "path/to/output"
)
```

### Panel Mode

Panel mode analyzes a specific set of genes across all samples:

```{r eval=FALSE}
cortar(
  file = "path/to/samplefile.tsv",
  mode = "panel",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "path/to/output",
  genelist = c("DMD", "TTN", "CFTR", "NF1")
)
```

### Research Mode

Research mode is similar to panel mode but with different filtering criteria:

```{r eval=FALSE}
cortar(
  file = "path/to/samplefile.tsv",
  mode = "research",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "path/to/output",
  genelist = c("GENE1", "GENE2")
)
```

## Understanding Parameters

### Required Parameters

- **file**: Path to your sample file
- **mode**: Analysis mode ("default", "panel", or "research")
- **assembly**: Genome assembly used for alignment ("hg38" or "hg19")
- **annotation**: Annotation source ("UCSC" or "1000genomes")
- **paired**: Whether RNA-seq is paired-end (TRUE or FALSE)
- **stranded**: Strandedness of RNA-seq (0 = unstranded, 1 = forward, 2 = reverse)
- **output_dir**: Directory for output files

### Optional Parameters

- **genelist**: Character vector of genes (required for panel and research modes)
- **prefix**: Prefix to add to output file names
- **debug**: Enable debug mode for troubleshooting (TRUE or FALSE)

## Batch Processing

You can run multiple sample files in sequence:

```{r eval=FALSE}
cortar_batch(
  folder = "path/to/samplefiles",
  pattern = "*.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "path/to/output"
)
```

## Output Files

Cortar generates the following outputs:

1. **Excel Report**: Contains detailed splicing event information with highlighting
2. **PDF Report**: Visual representation of splicing patterns
3. **Debug Files** (if enabled): Intermediate files for troubleshooting

## Troubleshooting

### Common Issues

**Issue**: "File does not exist or is not readable"
- **Solution**: Check that all BAM file paths in your sample file are correct and accessible

**Issue**: "Gene name is invalid"
- **Solution**: Verify that gene symbols are correct and exist in the selected assembly

**Issue**: "Parameter 'mode' has invalid value"
- **Solution**: Ensure mode is one of: "default", "panel", or "research"

### Debug Mode

Enable debug mode for detailed troubleshooting:

```{r eval=FALSE}
cortar(
  file = "path/to/samplefile.tsv",
  mode = "default",
  assembly = "hg38",
  annotation = "UCSC",
  paired = TRUE,
  stranded = 2,
  output_dir = "path/to/output",
  debug = TRUE
)
```

Debug files will be saved to `<output_dir>/debug/`

## Next Steps

- See the "Advanced Usage" vignette for complex scenarios
- Check the function documentation with `?cortar`
- Visit the GitHub repository for updates and issue reporting

## Session Info

```{r eval=FALSE}
sessionInfo()
```
